---
title: "S&DS 220 Chapter 1 code examples"
author: "Braeden Cullen"
date: "2024-01-16"
output:
  html_document:
    highlight: default
    toc: true
    number_sections: true
    toc_depth: 3
    theme: spacelab
    toc_float: true
    collapsed: true
    smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data in R

This document contains code examples of the topics contained in Chapter 1 of [*Probability, Statistics, and Data: A fresh approach using R*](https://mathstat.slu.edu/~speegle/_book/preface.html) by Darrin Speegle and Bryan Clair.

For each bullet point in the sections below, add a code chunk with code examples.  To add a code chunk, go to "Code -> Insert Chunk" or use the shortcut `option` + `command` + `i`.

```{r}
2 + 2
```

## Arithmetic and variable assignment

- operations: `+`, `-`, `*`, `/`, also `^`, `**`, `%%`
- exponentials and logarithms: `exp`, `log`
- storing variables `<-` and the `str` function
- comments `#`

## Help

- help with `?` and the "Help" tab in RStudio

## Vectors

- don't forget that R is 1-indexed, unlike most other languages like python

- the combine function `c -> used to create vectors, since we are combining entries
```{r}
c(1:4, -10:-14) # combines 1 to 4 and -10 to -14

```

- the `:` operator, makes a vector itself but can also be used c(1:10) in the same manner to create a vector
```{r}
3:10
```

- the `rep` function with arguments `times`, `length.out`, or `each`
```{r}
rep(c(1:4, -10:-14), times = 3)

rep(pi, times =)
```

- example with `c` and `rep`
```{r}
v1 <- rep(c("abe", "bello"), each = 5)
v2 <- rep(c("abe", "bello"), times = 5)
v1
v2
```
```{r}
v3 <- rep(1:52, length.out = 100)
v3
v4 <- rep(1:50, length.out = 100)
v4
```

- `seq` function with arguments `from`, `to` and `by`, or `length.out`
```{r}
seq(from = 0, to = 100, by = 1)
seq(from = 0, to = 1, by = 0.01)
seq(from = 0, to = 1, length.out = 11)
```

- operations on vectors
```{r}
x <- c(1, 2, 3, 4, 5)
2*x
x+2
x^2
```

- `sum`, `mean`, `max`, `min`
```{r}
sum(x)
mean(x)
max(x)
min(x)
```

- basic scatterplot `plot(x, y)`
```{r}
y <- c(2, 3, 4, 5, 6)
plot(x, y)

years <- 1900:2024
set.seed(10)
values <- rnorm(n = length(years), mean = 10, sd = 4)
plot(x = years, y = values, col = "salmon", type = "l")
plot(x = years, y = values, col = "salmon")
```
```{r}
hist(values, col="azure", breaks = seq(from = min(values), 
                                       to = max(values), 
                                       length.out = 15))
```
```{r}
plot(x = years,
     y = values,
     col = "salmon3",
     xlab = "Years",
     ylab = "Meaningless made up values",
     main = "Title")
```

## Indexing vectors

- indexing with numbers
```{r}
values[1:5]
values[10]
values[seq(from = 1, to = length(values), by = 2)]

```

- indexing with logicals
```{r}
which.max(values) # location of max val
which.min(values) # location of min val 
```

- indexing with comparisons: `==`, `!=`, `>`, `<`, `>=`, `<=`
```{r}
values[values > 15 | values < 5]
values > 10
values[!(values > 5)] # will give you all values !(>5)
```

- subset with `%in%`
- built in data sets: `rivers`, `discoveries`
- `head`, `sort`, `table`

```{r}
head(rivers, n = 10)
head(sort(rivers), n = 10)
head(sort(rivers, decreasing = TRUE), n=10)
head(sort(rivers, decreasing = FALSE, n=10))
```

## Data types

- numeric
- integer
- character
- logical
- factor
- complex
- raw
- missing values: `NA`, `na.rm = TRUE`

- note, NA is not true or false ! it is unknown

```{r}
exam_scores <- c(92, 70, 30, 42, 99, 88, 72, NA, 13)
exam_scores > 75
# mean(exam_scores) -> returns NA since can't take mean of vector which has NA
length(exam_scores)
mean(exam_scores, na.rm = TRUE)
```

## Data frames

- examples: `trees`, `mtcars`, `airquality`
- `str`, `head`, `summary`
- subsetting `[<rows> , <columns>]`
- rows/columns by numeric vectors, logicals
- columns by character vector with column names
- getting columns with `<dataset name>$<column name>`
- `which.max`, `which.min`

## Reading data from files

- `read.csv`
- working directory: `getwd()`, `setwd()`

## Packages

- `install.packages("<package name>")`
- `library(<package name>)`
