---
title: "S&DS 220 Chapter 1 code examples"
author: "Braeden Cullen"
date: "2024-01-16"
output:
  html_document:
    highlight: default
    toc: true
    number_sections: true
    toc_depth: 3
    theme: spacelab
    toc_float: true
    collapsed: true
    smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data in R

This document contains code examples of the topics contained in Chapter 1 of [*Probability, Statistics, and Data: A fresh approach using R*](https://mathstat.slu.edu/~speegle/_book/preface.html) by Darrin Speegle and Bryan Clair.

For each bullet point in the sections below, add a code chunk with code examples.  To add a code chunk, go to "Code -> Insert Chunk" or use the shortcut `option` + `command` + `i`.

```{r}
2 + 2
```

## Arithmetic and variable assignment

- operations: `+`, `-`, `*`, `/`, also `^`, `**`, `%%`
- exponentials and logarithms: `exp`, `log`
- storing variables `<-` and the `str` function
- comments `#`

## Help

- help with `?` and the "Help" tab in RStudio

## Vectors

- the combine function `c()`
- the `:` operator
- the `rep` function with arguments `times`, `length.out`, or `each`
- example with `c` and `rep`
- `seq` function with arguments `from`, `to` and `by`, or `length.out`
- operations on vectors
- `sum`, `mean`, `max`, `min`
- basic scatterplot `plot(x, y)`

## Indexing vectors

- indexing with numbers
- indexing with logicals
- indexing with comparisons: `==`, `!=`, `>`, `<`, `>=`, `<=`
- subset with `%in%`
- built in data sets: `rivers`, `discoveries`
- `head`, `sort`, `table`

## Data types

- numeric
- integer
- character
- logical
- factor
- complex
- raw
- missing values: `NA`, `na.rm = TRUE`

## Data frames

- examples: `trees`, `mtcars`, `airquality`
- `str`, `head`, `summary`
- subsetting `[<rows> , <columns>]`
- rows/columns by numeric vectors, logicals
- columns by character vector with column names
- getting columns with `<dataset name>$<column name>`
- `which.max`, `which.min`

## Reading data from files

- `read.csv`
- working directory: `getwd()`, `setwd()`

## Packages

- `install.packages("<package name>")`
- `library(<package name>)`
