---
title: "S&DS 220: Homework 4"
subtitle: "Due Friday February 9"
author: "<insert you name>"
output:
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load R packages used here
library(fosdata)
```

## Instructions

1. Complete the questions below.  Upload your knitted PDF solutions to Gradescope by the due date.
2. Your solutions should be a combination of writing and R code.  When writing, use complete sentences.
3. Previous homework assignments already had code chunks created for you.  Now it is up to you to insert R code chunks within each problem as needed.
4. You should aim for clear and concise communication (in both words and R code).

\newpage

## Problem set questions

### Question 1: Exercise 2.23

Suppose you do an experiment where you select ten people at random and ask their birthdays.

Here are three events:

- $A$: all ten people were born in February.
- $B$: the first person was born in February.
- $C$: the second person was born in January.

(a) Which pair(s) of these events are disjoint, if any?

*Solution.*

(b) Which pair(s) of these events are independent, if any?

*Solution.*


(c) What is $P(B | A)$?

*Solution.*


### Question 2: Exercise 2.26

Suppose a die is tossed three times. Let $A$ be the event "the first toss is a 5". Let $B$ be the event "the first toss is the largest number rolled" (the "largest" can be a tie). Determine, via simulation or otherwise, whether $A$ and $B$ are independent.

*Solution.*


### Question 3: Exercise 2.30

Suppose there is a new test that detects whether people have a disease. If a person has the disease, then the test correctly identifies that person as being sick 99.9\% of the time (*sensitivity* of the test). If a person does not have the disease, then the test correctly identifies the person as being well 97\% of the time (*specificity* of the test). Suppose that 2\% of the population has the disease. Find the probability that a randomly selected person has the disease given that they test positive for the disease.

*Solution.*


### Question 4: Exercise 2.35

You should answer these questions without using simulation (however, feel free to check your answer with simulation). Six standard six-sided dice are rolled.

(a) How many outcomes are there?

*Solution.*


(b) How many outcomes are there such that all of the dice are different numbers?

*Solution.*


(c) What is the probability that you obtain six different numbers when you roll six dice?

*Solution.*


\newpage

### Question 5: Loops in R

Read the Vignette: Loops in R at the end of chapter 3 on pages 82 and 83 in our text. Below is an example of a `while` loop that counts the number of coin tosses until the 50th time heads appears, where the probability of heads is 0.4.

```{r, eval=FALSE}
# example of a while loop

heads_50 <- replicate(1e4, {
  # initialize the number of heads to 0
  n_heads <- 0
  # initialize the number of coin tosses to 0
  n_tosses <- 0
  
  while(n_heads < 50){
    # simulate a coin toss
    coin_toss <- sample(c("H", "T"), size = 1, prob = c(0.4, 0.6))
    
    # increment the number of tosses by 1
    n_tosses <- n_tosses + 1
    
    # if the coin toss is heads, increment the number of heads by 1
    if(coin_toss == "H") {
      n_heads <- n_heads + 1
    }
  
  }
  # return the total number of tosses
  n_tosses
})
```

(a) Write a `while` loop to count the number of times a standard six-sided die is rolled until an accumulated total of 250 or more is achieved. Include your `while` loop code inside `replicate` to simulate this experiment 10,000 times.

*Solution.*


Use the results of the simulation you wrote in part (a) to answer parts (b) through (e). In other words, don't re-simulate the experiment.

(b) Estimate the mean and standard deviation of the number of rolls.

*Solution.*


(c) Make a histogram of the number of rolls. Give the histogram an appropriate title and axes labels.

*Solution.*


(d) Estimate the probability that 75 or fewer rolls are required.

*Solution.*


(e) Suppose after 70 rolls, the accumulated dice total less than 250. What is the probability an accumulated total of 250 or more will be achieved in the next 5 rolls?

*Solution.*

